\section{Conclusions}\label{sec:conclusions}

This paper has discussed different approaches for representing defeasibility
as used in law texts, by annotating rules with modifiers that explicate their
relation to other rules. We have notably presented encodings in classical
logic (\secrefs{sec:restr_precond}) and explained
how they are related. Quite a different approach,
based on Answer Set Programming, is presented in
\secref{sec:defeasible_asp}. All the encodings are motivated by the need to
explore the implementation of various forms of defeasibility in logics for
which well developed and powerful solvers such as SMT solvers and ASP solvers
already exist. The ASP based and SMT based approaches are complimentary to
each other. ASP's closed-world-assumption is useful when a legal verdict must
be reached with potentially incomplete information, when the truth value of
every atom is not explicitly known. For example in the speed limit example,
unless it is known explicitly that the car in question was a sports car, the
ASP solver would always return a speed limit lower than $320$. The SMT solver
would generate two models, one in which the car is a sports car and one in
which it is not. This could lead to two different speed limits, hence
requiring human intervention to determine the true speed limit in the scenario
being considered. On the other hand, when one does model checking or wants to
reason about meta-properties of the rules such as soundness of the rule-set,
the SMT solver approach is more suited than the ASP approach. Intuitively,
here we want to reason over all possible scenarios rather than restricting the
reasoning to a particular scenario being considered.

However, we are still at the beginning of the journey. To allow the two
approaches to work together coherently and seamlessly, a theoretical
comparison of the classical and ASP semantics presented here still has to be
carried out, and it has to be propped up by an empirical evaluation. For this
purpose, we are currently in the process of coding some real-life law texts in
L4, such as Singapore's Personal Data Protection
Act\footnote{\url{https://sso.agc.gov.sg/Act/PDPA2012}}.  An implementation of
the L4 ecosystem is under
way\footnote{\url{https://github.com/smucclaw/baby-l4}}, providing a
transpilation of L4 rules to both the SMT and the ASP world. The interaction
with SMT solvers is done through an SMT-LIB \cite{BarFT_SMTLIB}
interface. Advanced solvers, such as Z3 \cite{demoura_bjorner_z3_2008},
provide good support for quantification.

% , but it has not yet
% reached a stable, user-friendly status. It is implemented in Haskell and features
% an IDE based on VS Code and natural language processing via Grammatical
% Framework \citep{ranta_grammatical_2004}. Currently, only the coding of
% \secrefs{sec:restr_precond} has been implemented---the coding of
% \secref{sec:restr_deriv} would require a much more profound transformation of
% rules and assertions. Likewise, the ASP coding of \secref{sec:defeasible_asp}
% has only been done manually. The interaction with SMT solvers is done through
% an SMT-LIB \citep{BarFT_SMTLIB} interface, thus opening the possibility to
% interact with a wide range of solvers. As our rules typically contain
% quantification, reasoning with quantifiers is crucial, and the best support
% currently seems to be provided by Z3 \citep{demoura_bjorner_z3_2008}.



% We are
% fully aware of shortcomings of the current L4, which will strongly evolve in
% the next months, to include reasoning about deontics and about temporal
% relations. Integrating these aspects will not be easy, and this is one reason
% for not committing prematurely to one particular logical framework.


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
