\documentclass[]{ceurart}
%%
%% One can fix some overfulls
\sloppy


\input{settings}
\input{defs}
\newtheorem{example}{Example}
\newtheorem{proposition}{Proposition}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{multirow}


%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{listings}
%% auto break lines
\lstset{breaklines=true}

%%
%% end of the preamble, start of the body of the document source.

\begin{document}


%% Rights management information.
%% CC-BY is default license.
\copyrightyear{2022}
\copyrightclause{Copyright for this paper by its authors.
  Use permitted under Creative Commons License Attribution 4.0
  International (CC BY 4.0).}

%% This command is for the conference information
\conference{2nd Workshop on Goal-directed Execution of Answer Set
  Programs (GDE'22), August 1, 2022}

\title{Automating Defeasible Reasoning in Law with Answer Set Programming}

\author{Lim How Khang}[%
style=chinese,
orcid=0000-0002-9333-1364
%email=name@org.com,
%url=http://www. ,
]
%
\author{Avishkar Mahajan}[%
orcid=0000-0002-9925-1533
%email=name@org.com,
]
%
\author{Martin Strecker}[%
orcid=0000-0001-9953-9871
%email=name@org.com,
]
%
\author{Meng Weng Wong}[%
orcid=0000-0003-0419-9443
%email=name@org.com,
]
\address{Singapore Management University}

\begin{abstract}
The paper studies defeasible reasoning in rule-based systems, in particular
about legal norms and contracts. We identify rule modifiers that specify how
rules interact and how they can be overridden. We then define rule
transformations that eliminate these modifiers, leading in the end to a
translation of rules to formulas. For reasoning with and about rules, we
contrast two approaches, one in a classical logic with SMT solvers, which is
only briefly sketched, and one using non-monotonic logic with Answer Set
Programming solvers, described in more detail.
\end{abstract}

\begin{keywords}
  Knowledge representation and reasoning
  \sep
  Argumentation and law
  \sep
  Computational Law
  \sep
  Defeasible reasoning
\end{keywords}

\maketitle




%----------------------------------------------------------------------

\input{introduction}

% file intentionally commented out, text now integrated into l4_language_app.tex
% \input{l4_language}

\input{reasoning_rules}

%\input{defeasible}

\input{asp}

\input{conclusions}

\begin{acknowledgments}
The contributions of the members of the L4 team to this common effort are
thankfully acknowledged, in particular of Jason Morris who contributed his
experience with Answer Set Programming; of Jacob Tan and Ruslan Khafizov who
have participated in discussions about its contents and commented on the
paper; and of Liyana Muthalib who has proof-read a previous version.

This research is supported by the National Research Foundation (NRF),
Singapore, under its Industry Alignment Fund –- Pre-Positioning Programme, as
the Research Programme in Computational Law. Any opinions, findings and
conclusions or recommendations expressed in this material are those of the
authors and do not reflect the views of National Research Foundation,
Singapore.
\end{acknowledgments}


%----------------------------------------------------------------------
\bibliography{gde22}


\end{document}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
